---
title: "AOL Data Mining"
output: pdf_document
date: "2023-06-15"
---

## Loading dataset
```{r}
data <- read.csv("merged_data_cleaned.csv")
```


```{r}
library(tidyverse)
library(plyr)
library(dplyr)
library(ggplot2)
library(corrplot)
attach(data)
```


## Data Exploration
```{r}
head(data)
```

```{r}
print(ncol(data))
print(nrow(data))
print(names(data))
```
```{r}
str(data)
```

```{r}
summary(data)
```

```{r}
unique(Country.of.Origin)
```

```{r}
unique(Quakers)
```

```{r}
colSums(is.na(data))
```
## Replacing missing values

```{r}
data$altitude_low_meters[is.na(data$altitude_low_meter)] <- mean(data$altitude_low_meter, na.rm = TRUE)
data$altitude_high_meters[is.na(data$altitude_high_meters)] <- mean(data$altitude_high_meters, na.rm = TRUE)
data$altitude_mean_meters[is.na(data$altitude_mean_meters)] <- mean(data$altitude_mean_meters, na.rm = TRUE)
data$Quakers[is.na(data$Quakers)]<-round(mean(data$Quakers, na.rm = TRUE))
```


```{r}
sum(is.na(data))
```

```{r}
country <- table(Country.of.Origin)%>%                
  as.data.frame() %>% 
  arrange(desc(Freq))
```

```{r}
country
```

```{r}
countries<-table(Country.of.Origin)
countries<-sort(countries,decreasing=TRUE)
```


```{r}
barplot(head(countries,10),ylim=c(0,300),
main = "Top 10 Arabica Producers",
xlab = "Countries",
ylab = "Freq",
col = "lightblue", las = 2, cex.names = 0.8
) + grid()
```

## Creating 'var' dataframe using 'Variety' column in 'data' dataset

```{r}
var <- table(Variety)%>%                
  as.data.frame() %>% 
  arrange(desc(Freq))
var
```

## Dropping unnamed variety from dataframe

```{r}
var<- var[-4,]
var
```

## Creating 'method' dataframe using 'Processing.Method' column in 'data' dataset

```{r}
method <- table(Processing.Method)%>% as.data.frame() %>% arrange(desc(Freq))
method
```

## Dropping unnamed method from dataframe

```{r}
method<-method[-3,]
method
```

## Creating 'color' dataframe using 'Color' column in 'data' dataset

```{r}
color <- table(Color)%>% as.data.frame() %>% arrange(desc(Freq))
color
```

## Dropping unfilled color type

```{r}
color<-color[-2,]
color
```

## Visualizing distributions of Country of Origin, Variety, Processing Methods, Color

```{r}
par(mfrow=c(2,2))
ggplot(head(country,10), aes(x= reorder(Country.of.Origin, Freq), y=Freq, fill = Country.of.Origin)) + geom_bar(stat = "identity") + coord_flip() + labs(title = "Top 10 Arabica Producers",
       x = "Countries",
       y = "Countries")  + scale_fill_hue(c=80)

ggplot(head(var,10), aes(x=reorder(Variety,-Freq),y=Freq, fill= "")) + geom_bar(stat = "identity") + 
  labs(Title="Top 10 Varieties", x="Variety",y="Freq") + scale_x_discrete(guide = guide_axis(angle = 90)) + scale_fill_manual(values=c("lightblue"))

ggplot(method, aes(x=reorder(Processing.Method,-Freq),y=Freq, fill= "")) + geom_bar(stat = "identity") + 
  labs(Title="Processing Methods", x="Method",y="Freq") + scale_x_discrete(guide = guide_axis(n.dodge = 2)) + scale_fill_manual(values=c("bisque4"))

ggplot(color, aes(x=reorder(Color,-Freq),y=Freq, fill= "")) + geom_bar(stat = "identity") + 
  labs(Title="Processing Methods", x="Method",y="Freq")  + scale_fill_manual(values=c("aquamarine3"))


```


## Checking data distribution of Total Cup Points

```{r}
# getting samples
set.seed(1234)
dplyr::sample_n(data, 10)
shapiro.test(data$Total.Cup.Points)

# p value < 0.05 -> data is not normally distributed
```

```{r}
library("car")
par(mfrow = c(1,2))
hist(Total.Cup.Points)
qqPlot(Total.Cup.Points)
```


## Finding outliers

```{r}
library(gridExtra)
boxplot(Total.Cup.Points, main = "Total Cup Points")
```

## Replacing outlier with mean

```{r}
Total.Cup.Points <- replace(Total.Cup.Points, Total.Cup.Points == 0, mean(Total.Cup.Points)) 
```

## Finding country with highest total cup points

```{r}
by_country <- arabica %>% group_by(Country.of.Origin)
by_country %>% filter(Total.Cup.Points == max(Total.Cup.Points))
```

```{r}
print(arrange(by_country, -Total.Cup.Points))
```


## Finding countries with highest measure quality values

```{r}
# selecting quality measures variables
arabica <- select(data, Country.of.Origin, 21:34, Category.Two.Defects)
arabica
```


```{r}
aroma<-by_country %>% filter(Aroma == max(Aroma))
aro<-arrange(aroma,-Aroma)
print(head(select(aro,Country.of.Origin,Aroma)),5)

flavor<-by_country %>% filter(Flavor == max(Flavor))
flv<-arrange(flavor,-Flavor)
print(head(select(flv,Country.of.Origin,Flavor)),5)

aftertaste<-by_country %>% filter(Aftertaste == max(Aftertaste))
aft<-arrange(aftertaste,-Aftertaste)
print(head(select(aft,Country.of.Origin,Aftertaste)),5)

acidity<-by_country %>% filter(Acidity == max(Acidity))
aci<-arrange(acidity,-Acidity)
print(head(select(aci,Country.of.Origin,Acidity)),5)

body<-by_country %>% filter(Body == max(Body))
bod<-arrange(body,-Body)
print(head(select(bod,Country.of.Origin,Body)),5)

balance<-by_country %>% filter(Balance == max(Balance))
bal<-arrange(balance,-Balance)
print(head(select(bal,Country.of.Origin,Balance)),5)

```

```{r}
uniformity<-by_country %>% filter(Uniformity == max(Uniformity))
uni<-arrange(uniformity,-Uniformity)
print(head(select(uni,Country.of.Origin,Uniformity)),5)

clean<-by_country %>% filter(Clean.Cup == max(Clean.Cup))
cl<-arrange(clean,-Clean.Cup)
print(head(select(cl,Country.of.Origin,Clean.Cup)),5)

sweet<-by_country %>% filter(Sweetness == max(Sweetness))
swt<-arrange(sweet,-Sweetness)
print(head(select(swt,Country.of.Origin,Sweetness)),5)

moisture<-by_country %>% filter(Moisture == max(Moisture))
mois<-arrange(moisture,-Moisture)
print(head(select(mois,Country.of.Origin,Moisture)),5)

defect1<-by_country %>% filter(Category.One.Defects == max(Category.One.Defects))
def1<-arrange(defect1,-Category.One.Defects)
print(head(select(def1,Country.of.Origin,Category.One.Defects)),5)

defect2<-by_country %>% filter(Category.Two.Defects == max(Category.Two.Defects))
def2<-arrange(defect2,-Category.Two.Defects)
print(head(select(def2,Country.of.Origin,Category.Two.Defects)),5)
```


```{r}
# selecting only ethiopia from data
ethiopia_arabica <- arabica %>% 
  filter(Country.of.Origin == "Ethiopia")
# sorting in descending order
ethiopia_arabica %>% arrange(desc(Total.Cup.Points))
```


```{r}
# Removing unused columns
ethiopia_arabica$Country.of.Origin <- NULL
```

```{r}
df_correlation <- as.data.frame(round(cor(ethiopia_arabica %>% dplyr::select(everything())), 3))
library(corrplot)
corr_graph <- cor(ethiopia_arabica)
corrplot(corr_graph)
```

```{r}
par(mfrow = c(2,3))
ggplot(ethiopia_arabica, aes(Total.Cup.Points, Aroma)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Aroma")
ggplot(ethiopia_arabica, aes(Total.Cup.Points, Flavor)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Flavor")
ggplot(ethiopia_arabica, aes(Total.Cup.Points, Aftertaste)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Aftertaste")
ggplot(ethiopia_arabica, aes(Total.Cup.Points, Acidity)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Acidity")
ggplot(ethiopia_arabica, aes(Total.Cup.Points, Body)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Body")
ggplot(ethiopia_arabica, aes(Total.Cup.Points, Balance)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Balance")
```


## Finding ethiopia's smallest valued variable among top 6 correlated with total cup points

```{r}
summary(ethiopia_arabica[1:6])
```
```{r}
eth_aroma<-round(mean(ethiopia_arabica$Aroma),3)
eth_flv<-round(mean(ethiopia_arabica$Flavor),3)
eth_aft<-round(mean(ethiopia_arabica$Aftertaste),3)
eth_aci<-round(mean(ethiopia_arabica$Acidity),3)
eth_bal<-round(mean(ethiopia_arabica$Balance),3)
eth_bod<-round(mean(ethiopia_arabica$Body),3)
```


```{r}
eth_df<-data.frame(
  Measures=c("Aroma","Flavor","Aftertaste","Acidity","Balance","Body"),
  Value=c(eth_aroma,eth_flv,eth_aft,eth_aci,eth_bal,eth_bod))
eth_df
```


```{r}
ggplot(eth_df, aes(x=reorder(Measures,-Value),y=Value, fill = Measures)) + geom_bar(stat = "identity") + 
  labs(Title="Ethiopia Arabica Quality Measures", x="Measures",y="Value") + coord_cartesian(ylim=c(7.8,8.1)) + scale_fill_manual(values = c("saddlebrown", "tan", "saddlebrown", "saddlebrown", "saddlebrown", "saddlebrown"))
```

## Looking for variables that affect 'aftertaste'

```{r}
arabica2<-select(data,Harvest.Year,Variety:Category.Two.Defects,42:44)
ethiopia_arabica2 <- arabica2 %>% 
  filter(Country.of.Origin == "Ethiopia")
```

```{r}
ethiopia_arabica2
```


## Printing amount of unavailable data out of 44

```{r}
print(sum(ethiopia_arabica2$Harvest.Year == ""))
print(sum(ethiopia_arabica2$Variety == ""))
print(sum(ethiopia_arabica2$Processing.Method == ""))
print(sum(ethiopia_arabica2$Color == ""))
```
## Dropping unusable columns

```{r}
ethiopia_arabica2$Variety <- NULL
ethiopia_arabica2$Processing.Method <- NULL
ethiopia_arabica2$Color <- NULL
```

## Transforming chr data to int

```{r}
table(ethiopia_arabica2$Harvest.Year)
```
```{r}
# Removing vague data
ethiopia_arabica2<-subset(ethiopia_arabica2, Harvest.Year!="")
ethiopia_arabica2<-subset(ethiopia_arabica2, Harvest.Year!="August to December")
```


```{r}
ethiopia_arabica2<-transform(ethiopia_arabica2, Harvest.Year = as.integer(as.character(
       factor(Harvest.Year, 
         levels = c("2009/2010", "2010", "2012","2013","2014","2015","2016","2017","March 2010"),
         labels = c(1, 2, 3, 4, 5, 6, 7, 8, 2)
       )
 )))
```


## Plotting correlation with aftertaste

```{r}
coor_eth<-round(cor(ethiopia_arabica2),3)
coor_eth
```

```{r}
coreth <- as.data.frame(round(cor(ethiopia_arabica2 %>% dplyr::select(everything())), 3))
corr_graph <- cor(ethiopia_arabica2)
corrplot(corr_graph)
```
```{r}
eth_temp<-select(ethiopia_arabica2,Aroma:Uniformity,Sweetness,Cupper.Points,Quakers,Category.One.Defects,Category.Two.Defects,altitude_high_meters,altitude_low_meters,altitude_mean_meters)
eth_temp
```

```{r}
coreth_temp <- as.data.frame(round(cor(eth_temp %>% dplyr::select(everything())), 3))
corr_graph <- cor(eth_temp)
corrplot(corr_graph)
```

```{r}
par(mfrow = c(2,7))
ggplot(eth_temp, aes(Aftertaste, Aroma)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Aroma")
ggplot(eth_temp, aes(Aftertaste, Flavor)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Flavor")
ggplot(eth_temp, aes(Aftertaste, Acidity)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Acidity")
ggplot(eth_temp, aes(Aftertaste, Body)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Body")
ggplot(eth_temp, aes(Aftertaste, Balance)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Balance")
ggplot(eth_temp, aes(Aftertaste, Uniformity)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Uniformity")
ggplot(eth_temp, aes(Aftertaste, Sweetness)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Sweetness")
ggplot(eth_temp, aes(Aftertaste, Cupper.Points)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Cupper.Points")
ggplot(eth_temp, aes(Aftertaste, Quakers)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Quakers")
ggplot(eth_temp, aes(Aftertaste, Category.One.Defects)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Category.One.Defects")
ggplot(eth_temp, aes(Aftertaste, Category.Two.Defects)) + geom_point() + geom_smooth(method="lm") + labs(title="vs Category.Two.Defects")
ggplot(eth_temp, aes(Aftertaste, altitude_high_meters)) + geom_point() + geom_smooth(method="lm") + labs(title="vs altitude_high_meters")
ggplot(eth_temp, aes(Aftertaste, altitude_low_meters)) + geom_point() + geom_smooth(method="lm") + labs(title="vs altitude_low_meters")
ggplot(eth_temp, aes(Aftertaste, altitude_mean_meters)) + geom_point() + geom_smooth(method="lm") + labs(title="vs altitude_mean_meters")

```


## Removing irrelevant variables

```{r}
ethiopia_arabica2$Harvest.Year <- NULL
ethiopia_arabica2$Category.One.Defects <- NULL
ethiopia_arabica2$Quakers <- NULL
ethiopia_arabica2$Category.Two.Defects <- NULL
```


```{r}
ethiopia_arabica2$Cupper.Points <- NULL
ethiopia_arabica2$Total.Cup.Points <- NULL
```


## Predictive Modelling

```{r}
# splitting data into train and test data
set.seed(123)
samp <- sample(nrow(ethiopia_arabica2), 0.8 * nrow(ethiopia_arabica2))
train <- ethiopia_arabica2[samp, ]
test <- ethiopia_arabica2[-samp, ]
```

```{r}
print(dim(test))
print(dim(train))
```
```{r}
print(test)
print(train)
```

## Random Forest Model

```{r}
library(randomForest)
model <- randomForest(Aftertaste ~ . , data = train, mtree=1000, mtry=5)
model
```

```{r}
# number of trees that produce lowest test MSE
print(which.min(model$mse))
# find RMSE of best model
print(sqrt(model$mse[which.min(model$mse)]))
```
```{r}
# test MSE by number of trees
plot(model)
```

```{r}
# Variable importance plot
varImpPlot(model)
```


```{r}
# Model prediction results
prediction <- predict(model, newdata = test)
table(prediction,test$Aftertaste)
results<-cbind(prediction,test$Aftertaste)
colnames(results)<-c('pred','real')
results<-as.data.frame(results)

results
```
```{r}
library(Metrics)
mae(results$pred,results$real)
rmse(results$pred,results$real)
mse(results$pred,results$real)

actual <- results$real
predicted <- results$pred

ss_total <- sum((actual - mean(actual))^2)  # Total sum of squares
ss_residual <- sum((actual - predicted)^2)   # Residual sum of squares

r2_manual <- 1 - (ss_residual / ss_total)
print(r2_manual)

```

```{r}
# Calculating RMSE
sqrt(mean((results$real - results$pred)^2))
```

## Multi Linear Regression

```{r}
lm_model<-lm(Aftertaste~Balance + Flavor + Acidity + Body + Aroma + altitude_mean_meters + altitude_low_meters + altitude_high_meters + Moisture + Uniformity + Sweetness + Clean.Cup, data=ethiopia_arabica2)
summary(lm_model)
```
```{r}
pred_lm<-predict(lm_model,newdata=test)
table(pred_lm,test$Aftertaste)
results_lm<-cbind(pred_lm,test$Aftertaste)
colnames(results_lm)<-c('pred','real')
results_lm<-as.data.frame(results_lm)

results_lm
```
```{r}
sqrt(mean((results_lm$real - results_lm$pred)^2))
```
